---
title: Solutions GLM
author: B Kleinberg
date: 29 January 2019
subtitle: Dept of Security and Crime Science, UCL
output: html_notebook
---

*SOLUTIONS*

---

Tutorial 2, Probability, Statistics and Modelling 2, BSc Security and Crime Science, UCL

---

## Outcomes of this tutorial

In this tutorial, you will:

- build and evaluate your own multiple regression models
- build and evaluate your own logistic regression models
- conduct hypothesis testing
    - t-test
    - ANOVA
- perform generalized ANOVA/t-test testing

---

## Linear regression

Read the csv file located at `data/tutorial2/black_friday/sample_black_friday.csv` as a dataframe, and name it `blackfriday` [(Details on the data)](https://www.kaggle.com/mehdidag/black-friday/version/1).

```{r}
#your code
blackfriday = read.csv('data/tutorial2/black_friday/sample_black_friday.csv'
                       , header=T)
```

Familiarise yourself with the dataset:

```{r}
#your code
names(blackfriday)
head(blackfriday)
dim(blackfriday)
```

### Task 1:

Build a linear model that predicts the purchase size based on gender. Use the `glm` function to do this.

```{r}
#your code
lin_model = glm(Purchase ~ Gender
                , data=blackfriday
                , family = gaussian)
```

### Task 2:

Assess the model fit using a "fit" metric that tells you how far on average the model mis-estimated the purchase size.

```{r}
#your code
summary(lin_model)
```

```{r}
#your code
mean(abs(lin_model$residuals))
```


### Task 3:

Now build two competing models:

- one model with the added variable `Marital_Status`
- and one model with all terms possible from the two variables `Marital_Status` and `Gender`

```{r}
#your code
lin_model2 = glm(Purchase ~ Gender + Marital_Status
                , data=blackfriday
                , family = gaussian)

lin_model3 = glm(Purchase ~ Gender*Marital_Status
                , data=blackfriday
                , family = gaussian)
```

Now compare the sum of squared residuals of all three models:

```{r}
#your code
sum(lin_model$residuals^2)
sum(lin_model2$residuals^2)
sum(lin_model3$residuals^2)
```

Now compare the three models statistically. What are your conclusions?

_Hint:_ make sure to check the documentation for the ?anova function. You want to use the "F" test.

```{r}
#your code
anova(lin_model, lin_model2, test = 'F')
anova(lin_model2, lin_model3, test = 'F')
anova(lin_model, lin_model3, test = 'F')
```

## Logistic regression

Load the dataframe called `stop_search_met` from the file `data/tutorial2/stop_and_search/stop_and_search_sample.RData`.

```{r}
#your code
load('./data/tutorial2/stop_and_search/stop_and_search_sample.RData')
```

Familiarise yourself with the dataset:

```{r}
#your code
names(stop_search_met)
head(stop_search_met)
table(stop_search_met$Outcome)
```

### Task 4:

Subset the data to contain only the two levels of the variable `Outcome` "Arrest" and "A no further action disposal".

```{r}
#your code
stop_search_met_subset = stop_search_met[stop_search_met$Outcome == 'Arrest' | stop_search_met$Outcome == 'A no further action disposal', ]
```

Now build a logistic regression model on the `Outcome` variable as outcome variable modelled through the gender and age of the suspect.

```{r}
#your code
log_model = glm(Outcome ~ Gender + Age.range
                , data=stop_search_met_subset
                , family = binomial)
```

What are your conclusions about the effects of either predictor variable?

```{r}
#your code
summary(log_model)
```

```{r}
exp(coefficients(log_model))
```


### Task 5:

Expand the model and include the predictor variable `Officer.defined.ethnicity`.

```{r}
#your code
log_model2 = glm(Outcome ~ Gender + Age.range + Officer.defined.ethnicity
                , data=stop_search_met_subset
                , family = binomial)
```

How does this affect the model fit?

```{r}
#your code
summary(log_model2)
```

### Task 6:

Examine the variable `Type` of the original dataset:

```{r}
#your code
table(stop_search_met$Type)
```

Now exclude the level `Vehicle search` and build a logistic regression model on the two remaining levels. Try to answer the question whether race or gender affected the type of stop and search.

```{r}
#your code
stop_search_met_subset_2 = stop_search_met[stop_search_met$Type != 'Vehicle search', ]
```

```{r}
log_model3 = glm(Type ~ Gender+Officer.defined.ethnicity
                , data=stop_search_met_subset_2
                , family = binomial)
```

```{r}
summary(log_model3)
```

## GLM for group comparisons

Use the blackfriday dataset.

Check the documentation in R for the `aov(...)` function.

Note that an outstanding online help is [https://www.rdocumentation.org/](https://www.rdocumentation.org/).

### Task 7:

Does the amount of money spent on purchaes on black friday differ between female and male shoppers?

Show your results with a t-test, an ANOVA and the GLM.

```{r}
#your code
t.test(Purchase ~ Gender
       , data=blackfriday)
```


```{r}
#your code
summary(aov(Purchase ~ Gender
    , data=blackfriday))
```

```{r}
#your code
anova(glm(Purchase ~ Gender
    , data=blackfriday
    , family = gaussian)
    , test = 'F')
```


Show that 

Note that for the GLM implementation, you will have to run a model comparison against the model itself using the F test, to obtain the F-value.  
Note also that the `t.test` corrects for unequal variances, which the other two don't.

```{r}
#your code
(-4.24)^2
sqrt(17.981)
```

### Task 8:

Use an ANOVA and the GLM to test the hypothesis that age affects the purchase size.

```{r}
#your code
anova_1 = aov(Purchase ~ Age
              , data = blackfriday)

summary(anova_1)
```

```{r}
#your code
glm_1 = glm(Purchase ~ Age
            , data = blackfriday
            , family = gaussian)

summary(glm_1)
```

What is your conclusion?

### Task 9:

*Bonus question*

Use the `blackfriday` dataset and try to adopt the role of a business analyst. Suppose you are interested in customer profiling and want to target married customers differently than single customers.

Build a model that models the customers' marital status.

```{r}
#your code
```


## END

---

